{\bf Code Generation Log Replay for In-memory Database Management Systems}

### 2021-07-26 paper.tex
Updates from Overleaf

@@

% Provides a draft mark at the top of the document. 
%\draftstamp{\today}{DRAFT}

@@

\title{ %% {\it \huge Thesis Proposal}\\
{\bf Filter Representation in Vectorized Query Execution}}
\author{Amadou Latyr Ngom}
\date{July}
\Year{2020}
\trnumber{}

\committee{
Andy Pavlo, Chair \\
Todd C. Mowry \\
}

@@

\begin{abstract}
Advances in memory capacity have allowed Database Management Systems (DBMSs) to store large amounts of data in memory, thereby shifting the performance bottleneck of query execution from disk accesses to CPU efficiency (i.e., instruction count and cycles per instruction). One technique used to achieve such efficiency in analytical applications is batch-oriented processing or \textit{vectorization}: it reduces interpretation overhead, improves cache locality, and allows for efficient loop optimizations (e.g., loop unrolling, SIMD vectorization). For each \textit{vector} (i.e., a batch of tuples), the DBMS needs to track the set of valid tuples after a predicate application. To that end, systems employ one of two data structures, or \textit{filter representations}: Selection Vectors (SelVecs) or Bitmaps. In this work, we analyze each approach's strengths and weaknesses to provide recommendations on how to implement vectorized operations. Through a wide range of micro-benchmarks, we determine that the optimal implementation strategy is a function of many factors: the cost of iterating through tuples, the cost of the operation itself, and how amenable it is to SIMD vectorization. Our analysis shows that Bitmaps perform better for operations that can be efficiently vectorized using SIMD instructions but that SelVecs perform better on all other operations due to a cheaper iteration logic.
\end{abstract}

\begin{acknowledgments}
I would like to thank my advisor Andy Pavlo for his guidance and help during my last two years at CMU. He has developed my interest in databases and computer science research in general and made my experience at CMU very educational and fun. I would also like to thank my amazing peers of the CMUDB group that helped me in this and other projects: Prashanth Menon, Matt Buttrovich, Wan Shen Lim, Lin Ma, Tanuj Nayak, Rohan Aggarwal, Gustavo Angulo, Tianyu Li, John Rollinson. Special shoutout to Prashanth Menon, whose help was invaluable in this project. Finally, I would like to thank my parents and siblings who guided and supported their son/little bro all throughout his life.
\end{acknowledgments}

@@

\chapter{Introduction and Background}
\label{chapter1}
DBMSs typically handle two broad classes of applications: Online Transaction Processing (OLTP) and Online Analytical Processing (OLAP) applications. OLTP applications tend to be short-running, write-heavy queries that only touch a few entries in a database. OLAP applications are long-running, read-mostly queries that analyze large amounts of data. Due to the unique characteristics of each class, storage engines (i.e., the component of the DBMS stores the data), and query engines (i.e., the component that executes queries) must specialize storage formats and database algorithms to efficiently handle OLAP or OLTP applications \cite{one_size_fits_all}.

OLTP applications aim for fast access and update on individual entries. Therefore, to improve cache performance, OLTP-specialized storage engines \cite{mysql,postgres,aurora,voltdb}, use a row-store (i.e., a storage format that stores all the attributes of an entry contiguously in memory and in cache). OLAP applications, on the other hand, optimize for fast reads and computing on attributes across multiple entries (e.g., to find the average value of an attribute across all tuples). OLAP-specialized storage engines \cite{redshift,vertica,saphana,duckdb}, thus, use a column-store (i.e., a storage format that stores similar attributes of different entries contiguously). Though most traditional DBMSs \cite{postgres,mysql} focused on OLTP applications and thus used row stores, column stores have resulted in significant performance gains for OLAP applications \cite{abadi_phd,one_size_fits_all}.

Besides the storage engine, disk-accesses were, historically,  the performance bottleneck of OLAP applications. For example, it takes around $1ns$ to perform random access in the  L1 cache, $100ns$ for DRAM, and $17000ns$ for SSDs \cite{numbersbyyear}. Disk accesses, thus, dominate the execution time of any query whose data is primarily disk-resident. Recent advances in DRAM manufacturing have exponentially decreased the price and increased the capacity of main-memory. The cost of DRAM per GB was around \$1000 in 2000, \$10 in 2010, and less than \$1 in 2020 \cite{dramprices}. This dramatic decrease in cost, and the corresponding increase in capacity, have both obeyed Moore's Law \cite{moorelaw}. It is now possible to store large databases entirely in DRAM. Rather than optimize for disk access, the query engines of in-memory DBMSs (i.e., DBMSs whose data fits in memory) must optimize for CPU efficiency by reducing instruction count and cycles per instruction when executing queries.

\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/SampleQuery.png}
    \caption{\textbf{Sample SQL Query and Plan.}}
    \label{fig:intro_query}
\end{figure}

\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/Volcano.png}
    \caption{\textbf{The Iterator Model on the sample SQL query} -- The code in red indicates virtual function calls.}
    \label{fig:volcano_graph}
\end{figure}

The traditional query execution technique, the Iterator Model \cite{volcano}, is ill-suited for in-memory DBMSs due to CPU efficiency reasons. It executes a query by interpreting the query plan. In this model, each database operator implements a virtual \texttt{next()} function that returns the next tuple it can generate; each expression (i.e. addition, multiplication) also provides an virtual \texttt{evaluate()} function that takes in a tuple and returns a value. \cref{fig:volcano_graph} showcases the Iterator model with the sample query in \cref{fig:intro_query}. The operators \texttt{Scan A} and \texttt{Scan B} iterate through their respective tables and output the tuples to their parent operators. The filter only outputs tuples that satisfy the \texttt{WHERE} clause. The join first builds a hash table, then probes it to find matches. For every tuple -- of which there can be billions in an OLAP setting -- the virtual function calls (marked in red in the figure) constitute a significant overhead, which decreases CPU efficiency. Since the 2000s, there has been much work dedicated to optimizing the CPU efficiency of in-memory column stores for high-performance OLAP applications.

\citeauthor{monetdb2} proposed an alternative to the Iterator Model: the Materialization Model \cite{monetdb2}. In this model, rather than process and output one tuple at a time, each operator processes all its input tuples, then outputs all its results to the next operator. The authors also introduce specialized algorithms for each relational operator to exploit the columnar format of their system. Although the Materialization Model removes the Iterator Model's interpretation overhead, it can become memory-bound due to frequent materializations (e.g., copies) of large columns. To solve this issue, \citeauthor{boncz_vector} proposed a middle ground between the Materialization and Iterator Model: the Vectorization Model \cite{boncz_vector}. The thesis explores strategies for speeding up vectorized queries (i.e., queries executed using the Vectorization Model). As such, we will dedicate the rest of this chapter to an overview of the Vectorization Model. We discuss alternative query engine designs and optimizations for in-memory analytical DBMSs in \cref{chapter5}.

\section{The Vectorization Model}
\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/Vectorization.png}
    \caption{\textbf{The Vectorization Model with the Sample Query} -- The code is nearly identical to that in \cref{fig:volcano_graph}, except that it processes vectors.}
    \label{fig:vectorization_graph}
\end{figure}

The Vectorization Model is almost identical to the Iterator model except that it operates on batches of $T$ tuples or \textit{vectors} (e.g., 2048 tuples at a time). The \texttt{next()} function returns vectors and the \texttt{evaluate()} function operates on vectors too. \cref{fig:vectorization_graph} showcases Vectorization Model on the sample query in \cref{fig:intro_query}; it is near identical to \cref{fig:volcano_graph}. The difference is that processing occurs one vector at at time rather than one tuple at a time.

\begin{figure}[t!]
\centering
\begin{subfigure}[t]{.5\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/MulByConstSISD.png}
 \caption{Unoptimized version.}
  \label{fig:vector_mul_sisd}
\end{subfigure}%
\begin{subfigure}[t]{.5\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/MulByConstSIMD.png}
 \caption{Optimization version. There are two optimizations here: SIMD vectorization and loop unrolling with a factor of 2. The code in green indicates Intel's AVX512 instructions.}
  \label{fig:vector_mul_simd}
\end{subfigure}
\caption{\textbf{A Vectorized Multiplication Function} -- The programmer usually writes only the code on the left; the compiler can perform the optimizations on the right.}
\label{fig:vector_mul}
\end{figure}

This batch-oriented processing provides several benefits. First, virtual function calls occur for every $T$ tuples, which amortizes their overhead by a factor of $T$. Second, materializing $T$ tuples at a time instead of all tuples eliminates the memory pressure of the Materialization Model. Third, functions that operate on vectors, also known as \textit{vectorized primitives}, are usually amenable to loop optimizations (e.g., loop unrolling, SIMD vectorization) applied by either the compiler or the programmer. Consider, for a example, the primitive shown in \cref{fig:vector_mul_sisd}. It multiplies the elements in a vector by a constant number, and stores the result in a new vector. The loop of this primitive has independent iterations. Furthermore, the body of the loop only contains a load, a multiplication, and a store, which can all operate on SIMD vectors. As a result, the compiler or programmer can unroll and SIMD vectorize the loop. \cref{fig:vector_mul_simd} shows one way to optimize the loop. The body now contains AVX512 instructions \cite{intel2019} that perform that load, the multiplication, and the store, each operating on eight elements at a time. Besides, the loop unrolling (by a factor of two) optimization reduces the cost of loop maintenance.

Despite these benefits, the Vectorization Model introduces a new challenge. Individual tuples within a vector may be \textit{filtered out}, meaning that subsequent operators should not process them. For example, in \cref{fig:vectorization_graph}, the \texttt{Join} operator should not insert tuples that do no satisfy the \texttt{WHERE} clause in the \texttt{JoinHashTable}. It should only insert \textit{selected} tuples -- those not filtered out. Next, we discuss the methods used to identify selected tuples within a vector.

\section{Filter Representation}
\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.7]{images/FilterRepresentationIntro.png}
    \caption{\textbf{Filter Representation} -- Selection Vectors (left) index into selected tuples. Bitmaps (right) are only set on selected rows.}
    \label{fig:repr_intro}
\end{figure}

A \textit{filter representation} is the data structure used to identify selected tuples with a vector. \cref{fig:repr_intro} displays the two common representations: Selection Vectors (SelVecs) and Bitmaps. A SelVec stores a list of indices into the selected tuples. Tuples at an index stored within the SelVec are selected; tuples without an index are implicitly filtered out. A Bitmap, on the other hand, stores a list of bits. Its set bits indicate the selected tuples; its unset bits indicate filtered out ones. In the rest of this thesis, we will use the noun \textit{filter} to denote a particular instance of either data structure.

\subsection{Contribution}
Previous works on vectorized execution have chosen a filter representation strategy without providing an experimental comparison of the two approaches. Vectorwise \cite{vectorwise}, and works derived from it \cite{miro_adapt,everything_vectorized,sompolski_vec},  rely on SelVecs. IBM DB2's BLU \cite{ibm_blu} and the more recent VIP \cite{orestis_bitmap} rely on Bitmaps for the intermediary results of a table scan's \texttt{WHERE} clause filters, and SelVecs for other relational operators. Nonetheless, we find that supporting both representations, and dynamically choosing between them results in better performance than static implementations. Depending on the specific primitive and on the selectivity (i.e., the ratio of selected tuples) of its input vector, SelVecs can outperform Bitmaps and vice-versa. This work's main contribution is a methodology on how to optimize the performance of arbitrary vectorized primitives by taking into account filter representation, selectivity, and loop optimizations. We also provide recommendations for developers working on vectorized query engines on how to best implement primitives.

The remainder of this thesis is structured as follows. In \cref{chapter2}, we analyze the factors that influence the performance of vectorized primitive. We then derive insights for developers implementing vectorized query processing in \cref{chapter3}. Next, we use these insights on the OLAP queries of the TPCH \cite{tpch} benchmark and show the performance gains we obtain in \cref{chapter4}. Finally, we discuss related work in \cref{chapter5} and conclude in \cref{chapter6}.

\chapter{Computing on Filtered Vectors}
\label{chapter2}
The purpose of this chapter is to discuss the factors that influence the performance of vectorized primitives. We first categorize the kinds of operations that primitives perform and the possible ways to implement them. Afterward, we will present and explain the equation that quantifies the performance of each implementation. Finally, we raise the main questions whose answers, provided in \cref{chapter3}, will determine how to implement primitives optimally.

\section{Operations on Filtered Vectors}
\begin{figure}[t!]
\centering
\hspace*{\fill}%
\begin{subfigure}{.8\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/UpdateIntro.png}
 \caption{Update operation.}
  \label{fig:update_intro}
\end{subfigure}%
\hspace*{\fill}%
\vspace*{8pt}%

\hspace*{\fill}%

\begin{subfigure}{.8\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/MapIntro.png}
 \caption{Map operation.}
  \label{fig:map_intro}
\end{subfigure}
\caption{\textbf{Operations on Filtered Vectors} -- Updates change the set of selected tuple. Maps apply a function to the set of selected tuples without updating the set. Selected input tuples are in green. Selected output tuples are in blue.}
\label{fig:update_map_intro}
\end{figure}
\cref{fig:update_map_intro} shows the two kinds of primitives we seek to optimize: Updates and Maps. Updates apply a predicate function to the vector and modify the set of selected tuples accordingly. A scan filter (i.e., \texttt{WHERE} clause) is an example of an update operation. On the other hand, Maps keep the selected set but compute a new vector using a mapping function. A projection (i.e., \texttt{SELECT} clause) is an example of a map operation. There are other primitives, but these often have side-effects (e.g., insertion in a hash table for joins or an array for sorting) and are, therefore, not amenable to our optimizations for correctness reasons (e.g., Full Compute introduced in \cref{computestrategies}). As such, we focus on side-effect-free primitives.

\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.45]{images/NewJoinProbeExample.png}
    \caption{\textbf{Multi-Step Join Probe} -- The color green indicates selected items in the input vector. Orange indicates hash table lookup hits. Blue indicates exact matches after key comparison; these represent the output of the probe. Greyed out squares are filtered out tuples, or empty hash table chains.}
    \label{fig:probe_example}
\end{figure}
All relational operators are decomposable into a set of consecutive primitives. As mentioned above, simple relational operators like scan filters or projections consist of one individual primitive. Complex operators, like joins and aggregations, consist of multiple primitives. As a proof of concept, we showcase, in detail, the example of join probes in \cref{fig:probe_example}. The first primitive computes the hashes of the input tuples; the result is a vector of hash values. It is, therefore, a Map primitive. The second primitive performs a hash table lookup using the hashes to obtain a vector of hash table entries. It is also a Map primitive. Due to hashing collisions, the entries may not match the input, so the algorithm compares the join keys to filter out inexact matches. For example, key 90's hit in the hash table is an exact match, whereas key 73's is not. The third primitive performs this comparison; it is an Update primitive. The fourth primitive advances the hash table chains to obtain the next entries (e.g., 85 follows 37); it is a Map primitive. The probe then loops back to the comparison primitive to find exact matches (e.g., 85 now has an exact match). The set of exact matches constitutes the output of the join probe.

Other relational operators (e.g., aggregations) are similarly decomposable.

\section{Transitions}
\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{images/DisjunctionHowto.png}
    \caption{\textbf{Decomposed Disjunctions} -- We apply the first clause (in orange) on all input tuples. In the second step, however, we only apply the second clause (in red) on those tuples that do not pass the first clause. In the third step, we take the union between the two outputs. The output tuples are in blue.}
    \label{fig:disjunctive_filter_howto}
\end{figure}

In between each primitive, operators may need to perform a \textit{transition} (i.e., modifications to the filter passed into the next primitive). We now discuss transitions and their associated cost. Most of the time, this cost is non-existent because the same filter is passed on to the next primitive: the transition is \textit{implicit}. For example, the hashing primitive and the lookup primitive in \cref{fig:probe_example} have the same filter; the transition performs no modifications. On the other hand, the Chain Following primitive operates on entries that do not satisfy the comparison: the transition involves a set difference. Set unions are another type of transition. Decomposed disjunctions \cite{pcq} (shown in \cref{fig:disjunctive_filter_howto}), for example, involve both set differences and unions. They decompose disjunctive \texttt{WHERE} clauses into a sequence of primitives -- one each for each conditional term. Each primitive operates on only tuples that have not satisfied the previous conditional terms, hence the set differences. The final filter consists of tuples that satisfy at least one primitive, hence the unions.

To the best of our knowledge, transitions are either implicit or take the form of set differences (for elements that do not pass the previous predicate) and unions (to obtain the final result). These operations are efficient to implement with Bitmaps by leveraging single-cycle bitwise AVX512 instructions \cite{intel2019} that operate on 512 bits at a time. Efficiently implementing set operations on SelVecs is more complicated.  Whenever necessary, Updates can write to one additional SelVec for tuples that do not pass the filter; this second SelVec implicitly corresponds to the result of a set difference. Note that it does not overlap with the first SelVec, so a union of any of their subsets is just an array append, a fast operation that can leverage SIMD instructions.

To show that the overhead of these set operations is similar and negligible for both Bitmaps and SelVecs, we run an experiment that performs the decomposed disjunction shown in \cref{fig:disjunctive_filter_howto}. We synthetically generate the data so that each tuple has a $50\%$ chance of satisfying either conditional terms, the average case scenario. The data contains around $2000000$ tuples split into $1000$ vectors. The results are shown in \cref{fig:disjunctive_filter_perf}. The overheads are negligible; they are also nearly equal ($3.55\%$ and $4.06\%$ for Bitmaps and SelVecs, respectively). The representation, therefore, matters little when it comes to transitions. Given this, we will focus on optimizing primitives and will ignore transition costs.

\begin{figure}[t!]
    \centering
    \includegraphics[scale=0.5]{eval/transition.pdf}
    \caption{\textbf{Transition Overheads} -- The overheads are $3.55\%$ and $4.06\%$ for Bitmaps and SelVecs respectively.}
    \label{fig:disjunctive_filter_perf}
\end{figure}

\section{Compute Strategies}
\label{computestrategies}
\begin{figure}[t!]
\centering
\hspace*{\fill}%
\begin{subfigure}{.8\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/SelectiveCompute.png}
 \caption{Selective Compute. An \texttt{x} indicates an uninitialized value.}
  \label{fig:compute_intro_selective}
\end{subfigure}%
\hspace*{\fill}%
\vspace*{8pt}%

\hspace*{\fill}%

>>>>

\newcommand{\dbSQL}[1]{\texttt{\textbf{#1}}\xspace}

@@

% Provides a draft mark at the top of the document. 
\draftstamp{\today}{DRAFT}

@@

\title{ %% {\it \huge Thesis Proposal}\\
{\bf Code Generation Log Replay for In-memory Database Management Systems}}
\author{Tianlei Pan}
\date{June}
\Year{2021}
\trnumber{}

\committee{
Andy Pavlo, Chair \\
Wenting Ye \\
}

@@

\begin{abstract}
Instead of interpreting log records, we are extending our query code generation engine to create custom programs to replay log records. We will also be exploring for different optimization techniques to improve the recovery speed of recovery with code generation.
\end{abstract}

\begin{acknowledgments}
yes
\end{acknowledgments}

@@

\chapter{Introduction}
A \textit{database management system (DBMS)} is a software that is responsible for storing data, analyzing data, and interacting with applications. Depending on the use case, a DBMS can be designed to focus more on either capturing data, or analyzing data. \textit{On-line Analytical Processing (OLAP)} applications focus on reading, analyzing and aggregating cold data that is less likely to be modified. On the other hand, \textit{On-line Transaction Processing (OLTP)} applications work with write-heavy transactions that modify the database frequently.

Consequently, recovery and durability in OLTP applications are more of a pressing issue compared to OLAP systems. Naive or manual backups are usually sufficient for read-only OLAP applications. However, since OLTP databases are being updated by transactions constantly, naive backups will either cause the database to lose significant amount of data, or fail to preserve the \textit{ACID (atomicity, consistency, isolation, and durability)} properties. This requires OLTP applications to develop specialized recovery mechanism to handle these problems. Commercial DBMS often combines Write-ahead Logging (WAL) with checkpoints to preserve database durability.

In addition, OLTP systems also require specialized algorithms and implementations to execute write transactions. Query compilation is an important optimization technique used in OLTP systems to greatly reduce the number of instructions that need to be executed. A DBMS uses a processing model (e.g. Iterator, Vectorization, Materialization) to determine how the DBMS executes a query plan. The processing model is responsible for translating a query into an intermediate expression. The execution engine then breaks down the expression into various tasks for execution. To speed up this process, the DBMS can use an optimization technique known as \textit{Query Compilation}. Query compilation requires the DBMS to generate specialized code for execution tasks to increase execution throughput. These specialized code are often created using an off-shelf compiler into a C/C++ program that implements the functionality of a certain query.

While query compilation has proved successful in increasing the efficiency of a DBMS, it comes with several drawbacks. Implementing a query compilation system can be quite an ordeal. It requires additional knowledge of compiler systems (e.g. LLVM) and huge amounts of engineering work to translate different execution tasks. Moreover, the DBMS also need to modify other components to accommodate with an execution engine reliant on query compilation.

One of such components is the recovery system. Since the execution engine has adapted to executing compiled native code, the DBMS needs to implement an additional execution engine to interpret the log records. Instead, we propose a system design that will extend the query execution engine to create custom programs for replaying log records.

\chapter{Background}
\section{Database Recovery}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{images/Logging.png}
\caption{Write Ahead Logging.}
\label{fig:logging}
\end{figure}
Database recovery systems are responsible for preserving the ACID principles of a DBMS in case of failure. Prior to 1980s, early recovery systems rely heavily on non-volatile memory storage. Most of the DBMSs were disk-based and the recovery protocols assume the existence of dirty pages. System R\cite{system_r}, for instance, creates shadow copy of updated pages and maintains log record updates. Nowadays, both the capacity and reliability of main memory have increased significantly. This greatly improved the capabilities of in-memory DBMSs, but their performance is still be threatened by the need to synchronize non-volatile memory with persistent storage. Modern recovery systems aim to reduce the synchronization overhead.

Recovery systems consist of two components: logging and recovery. Logging is the action of storing transactional information on disk. These information are stored in a special data structure called a log record. A log record can either contain physical information (e.g. changes made to a specific physical address) or higher-level information (e.g. user-input query). In Write-ahead logging \cite{aries}, the DBMS records database changes in a log file before the changes can be persisted on disk. Each log record in the log file usually consists of redo records (install the effects of a committed transaction) and undo records (remove the effects on an incomplete or aborted transaction). Since in-memory DBMSs do not persist uncommitted changes, undo records are no longer required for the recovery system.

Recovery is the action of restoring the database system into a consistent state. If available, the DBMS first attempts to restore its previously recorded checkpoint. The database then initiates the log replay process to go through every log record stored since that checkpoint. SiloR\cite{silo_r} uses physical logging, checkpoints and an epoch-based recovery system. Transactions are assigned with different epochs and can only be persisted when their epochs are smaller then the current persistent epoch. VoltDB\cite{malviya14} uses a form of physiological logging that reduces transaction execution overhead. SAP HANA\cite{lee18}, however, uses a log-less approach through parallel table replication.

\section{Query Compilation / Code Generation}
When a query arrives, the DBMS parses the query into an expression tree. Eventually, the optimizer of the DBMS converts the expression tree into a physical plan tree. The physical plan represents how the query will be executed. It usually consists of operators that specify physical operations on the DBMS (e.g. inserting values into a certain memory location). For instance, \dbSQL{UPDATE x SET col1 = -col1 WHERE col1 BETWEEN 0 AND 100} can be represented by an update plan tree with a sequential plan as a child. The execution engine then uses the plan tree for execution.

The Iterator model\cite{volcano} is widely used in many disk-based DBMS (e.g. ORACLE RDBMS, MySQL, Microsft SQL Server) to interpret plan trees. The model "iterates" over each operator by implementing a \dbSQL{next()} function in each operator that returns the next tuples to be operated. The expressions inside the operators are evaluated using the provided tuples and generates output tuples. This approach suffers from frequent virtual function calls for in-memory DBMS. The Materialization model reduces this overhead by processing all the input tuples and emitting all the outputs at once ("materializing" the result as a single row or column). The Materialization model is suitable for OLTP DBMS since each OLTP query only accesses a small number of tuples at a time. However, since the Materialization results are stored in an in-memory buffer, the execution throughput is then bound by the size of the available memory space. While disk-based DBMSs are less concerned about the in-memory access since most of their operations are disk-based, this is especially critical for in-memory DBMSs.

Vectorization and Query Compilation are two state-of-the-art techniques employed for in-memory DBMSs to reduce memory overhead\cite{kersten18}. The Vectorization model is a variant of the Iterator model. Instead of emitting a single tuple at a time, the Vectorization model emits a batch of tuples at a time to amortize the cost of memory access. The functions that operate on these tuples are also specialized to be able to take in a batch of tuples at a time and more efficient compared to iterating over a single tuple at a time.

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{images/QueryCompilation.png}
\caption{Query Compilation architecture.}
\label{fig:query_compilation}
\end{figure}

Query Compilation eliminates the cost of interpretation by compiling the physical plan to machine code. The machine code of the corresponding physical plan can then be executed repeatedly by the execution engine, removing any need for virtual function calls. Before being compiled to machine code, the physical plan is often first converted into an intermediate representation (e.g. C/C++ code, LLVM IR, or custom Domain Specific Language) that follows the grammar of an imperative language. This simplifies the process of converting the physical plan into machine code, as the intermediate representation form is designed to resembles closely to SQL statements.

Hyper\cite{neumann11} is one of the pioneers of query compilation. It uses a data-centric compilation approach that splits the query plan into pipelines and compiles each pipeline separately. Vectorwise\cite{raducanu13} chooses to pre-compile huge amounts of micro function primitives that perform simple operations on typed data. During runtime, the DBMS can directly invoke these compiled primitives for query execution.

\chapter{Method}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{images/Proposal.png}
\caption{Code Generation Recovery Mechanism.}
\label{fig:codegen_recovery}
\end{figure}

Our proposal is to extend the recovery layer and query compilation layer to support the interpretation of log records. This approach can be realized in three steps:

\begin{enumerate}
\item Extract information from log records.
\item Convert the information into expressions that can be accepted and compiled by the code generation engine (Code Generation Step).
\item Run the compiled machine code using the execution engine.
\end{enumerate}

This code generation approach brings around several benefits. Firstly, it significantly reduces engineering overhead. Compiling contents of the log records into machine code allows the recovery manager to seamlessly integrate with the built-in execution engine. This removes the need to specifically construct code to handle tasks that are not compiled to machine code.

\begin{figure}[H]
\centering
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/RecoverIndex.png}
 \caption{Index Update During Recovery.}
  \label{fig:pipeline_graph}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/RecoverConstraint.png}
 \caption{Constraint Checking During Recovery.}
  \label{fig:pipeline_code}
\end{subfigure}
\caption{Problems faced by conventional recovery in code generation DBMS.}
\label{fig:log_record}
\end{figure}

Secondly, it also resolves tricky problems that will appear when writing specialized code for log records. An example of this is index updates. Without code generation, the recovery manager needs to specifically reconstruct the index given an insert or update record. With code generation, the index updates are taken care directly by the execution engine that already contains high-performance specialized code for index maintenance. Similar problems, such as constraint checking

Our proposed architecture is implemented and tested in \textit{Noisepage} \cite{noisepage}, a self-driving in-memory DBMS developed at Carnegie Mellon University. Noisepage is built in C++ and uses the PostgreSQL wire protocol for user communication. The DBMS supports ACID transactions and \texttt{SNAPSHOT-ISOLATION}. Noisepage uses the conventional WAL recovery scheme with physical logging and depends on query compilation for query execution. We will dive into the details of Noisepage's recovery structure in the following sections.

\section{Recovery System Architecture}
\subsection{Transaction}
Transactions are the atomic operations of any DBMS. Noisepage uses a transaction to perform physical updates on the database and physical logging to disk. Noisepage uses a multi-versioned delta store transaction engine that ensures \texttt{SNAPSHOT-ISOLATION}. The transaction engine allows non-blocking reads over writes and vice versa, but does not allow write-write conflicts on a per tuple basis. A transaction's lifecycle begins when the transaction begins, and ends after the log manager has serialized its changes to disk.

Each data tuple in noisepage is uniquely identified through a tuple slot that stores the offset of a tuple. When the DBMS inserts a tuple into a table, it creates a new tuple slot that points to a memory location within the table. Subsequent updates to the tuple does not create direct copies of the tuple, but stores delta information (redo records) about the updates instead. These redo records are essential for executing log replay correctly. The structure of a redo record is shown later in \ref{fig:log_record}.

A transaction uses a buffer (redo buffer) that stores all the delta records that will appear in its lifetime. Each redo buffer has a fixed size of 4096 bytes and is allocated from a centralized buffer pool. Each time a transaction needs to change the contents of the DBMS, it will attempt to request space from the redo buffer to dump a corresponding delta record of the change. If the redo buffer runs out of space, the transaction will replace the current buffer with a new buffer from the buffer pool. The redo buffer allows the logging components to process the changes before the transaction is completed.

When a transaction commits, the transaction engine creates a commit record that contains a timestamp that refers to the oldest active transaction from the timestamp manager. The commit record is appended to the redo buffer. The redo buffer is then carried over to the log manager, where the changes will be serialized to disk. When a transaction aborts, the transaction creates an abort record that will prevent corresponding records to be replayed during recovery. Abort records are essential; since the redo buffer is persisted to disk by the log manager once it is full, it is possible for aborted transactions to persist on disk. Aborted transactions that are persisted on disk are rare, however, since it would be difficult for a typical OLTP transaction to fill up the entire redo buffer.

\subsection{Write-Ahead Logging}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{images/WAL.png}
\caption{Logging Architecture. Log manager uses the log serializer and log consumers to convert redo buffers into log records.}
\label{fig:wal}
\end{figure}

The log manager is responsible for performing write-ahead logging in Noisepage. All the changes are serialized to disk as logs before they are performed on the DBMS storage. Each log record is self-contained and does not require any additional metadata to be replayed. The log manager consists of two separate tasks that live in different threads: the Log Serializer tasks and the Log Consumer tasks.

\subsubsection{Log Serializer}
The log serializer task receives redo buffers from a transaction once it is full, or when the transaction decides to commit. The log serializer breaks down the redo buffers into raw memory segmented into fixed-size buffers of 4096 bytes. These buffers will be consumed by the Disk Log Consumer for persistence.

The log records must be serialized in the correct order for recovery to perform correctly. Different transactions can share a same redo buffer that is filled with each of their own log records. Moreover, it is possible for transactions to appear in a non-serial order relative to their begin timestamp. However, the log records performed by an individual transaction is guaranteed to appear in the same order as they were created. This property helps ensure \texttt{SNAPSHOT-ISOLATION} property of Noisepage during log replay.
\subsubsection{Log Consumer}
Buffers supplied by the log serializer can be consumed by different log consumer tasks. Each log consumer is given a copy of the original buffer to achieve parallelism. Noisepage uses two type of log consumers: disk log consumer and replication log consumer.

The disk log consumer repeatedly polls for new buffers from the log serializer and writes the changes to disk. To improve performance, disk log consumer uses group commit that allows a batch of changes to be committed over a time period specified by the user. Under default settings, the disk log consumer writes down changes every 10 microseconds of if more than 1 MB of data has been written since the last write.

The replication log consumer also polls for new buffers, but instead of persisting them to disk, sends the new buffers over across the network to database replicas.

\subsection{Log Replay}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{images/RecoveryManager.png}
\caption{How recovery operates in Noisepage.}
\label{fig:recoverymanager}
\end{figure}

Noisepage uses four different record types: \textit{Redo, Delete, Commit, Abort}. Redo and Delete records are not processed immediately. Instead, they are placed inside a deferred changes buffer, ordered by their transaction timestamp. An Abort record removes the corresponding Redo/Delete record from the buffered changes. A Commit record initiates a transaction to iterate over the deferred changes buffer. Once all the deferred changes are processed, the buffer is cleared and the transaction is committed. The transaction aborts if any log record is malformed or any recovery operation fails.

\subsubsection{Data Structure}
\begin{figure}[H]
\centering
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/RedoRecord.png}
 \caption{Redo Record Structure.}
  \label{fig:pipeline_graph}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/DeleteRecord.png}
 \caption{Delete Record Structure.}
  \label{fig:pipeline_code}
\end{subfigure}
\caption{The internal log structures.}
\label{fig:log_record}
\end{figure}

A physical log record consists of a header followed by a record body. The header stores metadata about the log record itself, including its record type, size and begin timestamp of the transaction that generated the log record as mentioned in Section 3.1.1. The record body contains physical information that is necessary for recovery and varies depending on the record type. Depending on the value type, the values may bee stored in-line or not in-inline. Simple types such as \texttt{INTEGER}, \texttt{FLOAT} can be stored in-line, while \texttt{VARLEN} entry may stores the pointer in the values column that points to another location in the record.

Tuple slots in the log records will no longer be valid memory locations during recovery. Instead, the recovery manager creates an internal mapping from original tuple slots to a new tuple slots that will be retrieved after new values are inserted. The mapping allows the recovery manager to apply changes to the correct memory locations in the new memory environment.

Since only Redo and Delete records will effectively modify physical storage, we only need to analyze the record body of those two record types. Modifications of a query to the DBMS can be resolved down to the modification of tuple slots on the physical level.

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{images/Delta.png}
\caption{Redo Record Delta.}
\label{fig:logging}
\end{figure}

 While storing tuple slots is sufficient for delete records, Redo Records require additional data to perform Recovery Insert and Update operations. These data are kept in the Record Delta, which stores a mapping from column ids to values.

\subsubsection{Recovery Operation}
A Noisepage recovery step has three different operations:
\begin{itemize}
    \item \textbf{Insert}: Represented by a Redo Record. The recovery manager inserts the values stored within the redo record into the given tuple slot and updates the indexes.
    \item \textbf{Delete}: Represented by a Delete Record. The recovery manager deletes the tuple slot specified by the delete record and updates the indexes.
    \item \textbf{Update}: Represented by a Redo Record. The recovery manager updates the tuple slot using the values within the redo record and updates the indexes.
\end{itemize}

While both Insert and Update are represented using the same Redo Record structure, they can be differentiated by keeping track of all the tuple slots found in log records. If a tuple slot inside a Redo Record is unseen before, then the Redo Record represents an Insert operation. Otherwise, the Redo Record represents an Update operation.

\section{Recovery Code Generation}
The code generation structure in Noisepage follows from the data-centric compilation approach in \cite{neumann11}, as seen in \ref{fig:query_compilation}. A query is interpreted, optimized and converted into a physical plan tree. A plan tree consists of plan nodes and specifies how the query should be executed on the physical level of the database.
